On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-01 manuscripts/PNAS/chazthm_pnas.tex

%\begin{proposition}
%If $E$ is an $\ell$-regular hypergraph of rank $k$ satisfying the singleton intersection property then $|E| \geq \ell m/k$. 
%\end{proposition}
%<<<<<<< HEAD
%As mentioned above, it is easy to construct such $E$ with $|E| = m$, implying the lower bound for sample size $N$ from the introduction.
%A straightforward calculation verifies that if $E$ is an $\ell$-regular hypergraph of rank $k$ satisfying the SIP, then $|E| \geq \ell m/k$.
%%Note that if $E$ is an $\ell$-regular hypergraph of rank $k$ satisfying the SIP, then $|E| \geq \ell m/k$: \[|E|k \geq \sum_{S \in E}|S| = \sum_{i \in [m]} \deg(i) = \ell m.\] 
%In certain cases, there are simple constructions achieving this minimum; for example, when $k = \sqrt{m}$ one can take $E$ to be the rows and columns formed by arranging $[m]$ in a square grid.  In light of Cor.~\ref{DeterministicUniquenessCorollary}, it would be interesting to determine other minimal such $E$.
%=======
%Note that every $\ell$-regular hypergraph $E \subseteq {[m] \choose k}$ satisfies: \[|E|k = \sum_{S \in E}|S| = \sum_{i \in [m]} \deg(i) = \ell m.\] 
As mentioned above, it is easy to construct a regular hypergraph $E \subseteq {[m] \choose k}$ with the SIP having size $|E| = m$, implying the lower bound for sample size $N$ from the introduction. In certain cases, the SIP can be achieved with $|E| < m$; for example, when $k = \sqrt{m}$ one can take $E$ to be the rows and columns formed by arranging $[m]$ in a square grid.  In light of Cor.~\ref{DeterministicUniquenessCorollary}, it would be interesting to determine other such $E$. %[ ** TODO ** can we show for every $k, m$ that there is an $E$ satisfying the SIP with $|E| = \ell m/k$ for every $\ell$ such that $\ell m/k$ is an integer? Then lower bound on $N$ comes from the smallest such $\ell$.]

@@

\begin{theorem}
Fix $n, \ell$, and $k < m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} have at least \mbox{$k \left[ (k-1){ m \choose k} + 1 \right]$} vectors supported in general linear position on each set in an $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  Given any $\delta_1, \delta_2 \geq 0$ and $\varepsilon \leq \varepsilon(\delta_1, \delta_2)$, the following holds.  % (\eqref{epsdel}).
% $\varepsilon < L_2(\mathbf{A}) / C_1$.
%There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.
Every solution $\mathbf{B}, \mathbf{\hat x}_i, \ldots, \mathbf{\hat x}_N$ to the constrained $\ell_0$ optimization problem:
\begin{align}\label{minsum}
\arg \min \sum_{i = 1}^N \|\mathbf{\hat x}_i\|_0 \ \ \text{s.t.} \ \ \|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{$i \in [N]$},
\end{align}
necessarily satisfies the recovery inequalities in \eqref{def1}. % \eqref{Cstable} and \eqref{b-PDa}.
\end{theorem}

\begin{proof}
We derive a lower bound on the number of $k$-sparse  $\mathbf{\hat x}_i$ and then apply Thm.~\ref{DeterministicUniquenessCorollary}. 
[*** This proof still has a gap ***]
Let $n_p$ be the number of $\mathbf{\hat x}_i$ with $\|\mathbf{\hat x}_i\|_0 = p$.  %, so that $\sum_{p = 1}^{m} n_p = N$. 
Since the $\mathbf{x}_i$ are all $k$-sparse, by \eqref{minsum} we have:
\begin{align}
k \sum_{p = 1}^{m} n_p = kN \geq \sum_{i=1}^N \|\mathbf{x}_i\|_0 \geq \sum_{i=1}^N \|\mathbf{\hat x}_i\|_0 = \sum_{p=1}^{m} p n_p 
\end{align}
\[\implies \sum_{p = 1}^k (k-p)n_p \geq \sum_{p = k+1}^{m} (p-k) n_p.\]
Hence,
\begin{align}
(k-1) \sum_{p =1}^k n_p \geq \sum_{p = 1}^k (k-p)n_p \geq \sum_{p = k+1}^{m} (p-k) n_p
\end{align}
\[ \geq \sum_{p = k+1}^{m} n_p = N - \sum_{p = 1}^k n_p \implies \sum_{p =1}^k n_p \geq N / k.\]

In particular, at least $N / k$ of the $\mathbf{\hat x}_i$ are $k$-sparse. Therefore, there is some $N' \subseteq [N]$ for which for each $S \in E$ there are $(k-1){m \choose k}+1$ of the vectors $\mathbf{x}_i$ with $i \in [N']$ supported on $S$ and for which $\mathbf{\hat x}_i$ is $k$-sparse for all $i \in [N']$.  The result now follows directly from Thm.~\ref{DeterministicUniquenessCorollary}. % \eqref{Cstable} and \eqref{b-PDa} follow.
\end{proof}

>>>>

%\begin{proposition}
%If $E$ is an $\ell$-regular hypergraph of rank $k$ satisfying the singleton intersection property then $|E| \geq \ell m/k$. 
%\end{proposition}
%Note that every $\ell$-regular hypergraph $E \subseteq {[m] \choose k}$ satisfies: \[|E|k = \sum_{S \in E}|S| = \sum_{i \in [m]} \deg(i) = \ell m.\] 
As mentioned above, it is easy to construct a regular hypergraph $E \subseteq {[m] \choose k}$ with the SIP having size $|E| = m$, implying the lower bound for sample size $N$ from the introduction. In certain cases, the SIP can be achieved with $|E| < m$; for example, when $k = \sqrt{m}$ one can take $E$ to be the rows and columns formed by arranging $[m]$ in a square grid.  In light of Cor.~\ref{DeterministicUniquenessCorollary}, it would be interesting to determine other such $E$. %[ ** TODO ** can we show for every $k, m$ that there is an $E$ satisfying the SIP with $|E| = \ell m/k$ for every $\ell$ such that $\ell m/k$ is an integer? Then lower bound on $N$ comes from the smallest such $\ell$.]

@@

\begin{corollary}
Fix $n, \ell$, and $k < m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} have at least \mbox{$\left[ (k-1){ m \choose k} + 1 \right] + (k-1)^2{m \choose k-1}$} vectors supported in general linear position on each set in an $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP. Given any $\delta_1, \delta_2 \geq 0$ and $\varepsilon \leq \varepsilon(\delta_1, \delta_2)$, the following holds.  % (\eqref{epsdel}).
% $\varepsilon < L_2(\mathbf{A}) / C_1$.
%There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.
Every solution $\mathbf{B}, \mathbf{\hat x}_i, \ldots, \mathbf{\hat x}_N$ to the constrained $\ell_0$ optimization problem:
\begin{align}\label{minsum}
\arg \min \sum_{i = 1}^N \|\mathbf{\hat x}_i\|_0 \ \ \text{s.t.} \ \ \mathbf{B}\mathbf{\hat x}_i = \mathbf{A}\mathbf{x}_i, \ \ \text{$i \in [N]$}, %\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{$i \in [N]$},
\end{align}
necessarily satisfies the recovery inequalities in \eqref{def1}. % \eqref{Cstable} and \eqref{b-PDa}.
\end{corollary}

\begin{proof}
We derive a lower bound on the number of $k$-sparse $\mathbf{\hat x}_i$ and then apply Thm.~\ref{DeterministicUniquenessCorollary}. 
First, note that $\|\mathbf{\hat x}_i\|_0 \neq 0$ for all $i$, since otherwise we would have $\mathbf{A}\mathbf{x}_i = \mathbf{B}\mathbf{\hat x}_i = \mathbf{0} \implies \mathbf{x}_i = \mathbf{0}$ by the spark condition, contradicting the general linear position of the $\mathbf{x}_i$. Let $n_p$ be the number of $\mathbf{\hat x}_i$ with $\|\mathbf{\hat x}_i\|_0 = p$.  %, so that $\sum_{p = 1}^{m} n_p = N$. 
Since the $\mathbf{x}_i$ are all $k$-sparse, by \eqref{minsum} we have:
\begin{align}
k \sum_{p = 1}^{m} n_p = kN \geq \sum_{i=1}^N \|\mathbf{x}_i\|_0 \geq \sum_{i=1}^N \|\mathbf{\hat x}_i\|_0 = \sum_{p=1}^{m} p n_p 
\end{align}
\[\implies \sum_{p = 1}^k (k-p)n_p \geq \sum_{p = k+1}^{m} (p-k) n_p.\]
Hence,
\begin{align}
\sum_{p = k+1}^m n_p \leq \sum_{p = k+1}^m (p-k) n_p \leq \sum_{p = 1}^k (k-p)n_p \leq (k-1) \sum_{p = 1}^{k-1} n_p.
\end{align}
%\[ \implies \sum_{p = k+1}^m n_p \leq (k-1)^2 {\hat m \choose k-1} \]
By the spark condition and general linear position of the $\mathbf{x}_i$, every $k$ vectors $\mathbf{Ax}_i$ span a $k$-dimensional space; hence, at most $k-1$ vectors $\mathbf{\hat x}_i$ can share a support of size $k-1$ or less, and the number of $(k-1)$-sparse vectors $\mathbf{\hat x}_i$ is then bounded from above by $(k-1) { m \choose k-1}$. It follows that at least $N - (k-1)^2 {m \choose k-1}$ of the vectors $\mathbf{\hat  x}_i$ are $k$-sparse, and that there must then be some $N' \subseteq [N]$ for which for each $S \in E$ there are $(k-1){m \choose k}+1$ of the vectors $\mathbf{x}_i$ with $i \in [N']$ supported on $S$ and for which $\mathbf{\hat x}_i$ is $k$-sparse for all $i \in [N']$.  The result now follows directly from Thm.~\ref{DeterministicUniquenessCorollary}. % \eqref{Cstable} and \eqref{b-PDa} follow.
\end{proof}
